---
title: "Disproportionate exposure to urban heat island intensity in the USA"
author: Muhammad Rifqi Febrian (mrf2174), Agastia Cestyakara (ac4973)
date: December 6, 2022
output: 
  ioslides_presentation: default
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{css, echo = FALSE}
table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
}
```

```{r}
#load library
library(tidyverse)
library(forcats)
library(fastDummies)
library(SciViews)
library(weights)
library(lmtest)
library(sandwich)
library(Weighted.Desc.Stat)
library(kableExtra)
library(stargazer)
library(sf)
library(ggplot2)
library(tidyverse)
library(tmap)
library("spData")
library("rgdal")
library(patchwork)

#load data
load("uhindvitract.RData")
```

## Introduction {.build}
<div>
- Urban areas are usually warmer than the rural areas that surround them — Urban heat island
</div>
<div>
- It could occurs because cities consume huge amounts of energy in electricity and fuel, have less vegetation to provide shade and cooling, and are built of materials that absorb and store heat
<div>
- Heat exposure is associated with heat-related mortality, illness, and non-fatal health outcomes.
</div>
<div>
- Urban heat islands (UHI) might disproportionately impact vulnerable communities
</div>
<div>
- Well-targeted interventions are needed to adapt and mitigate to the UHI effect
</div>

## Research questions {.build}
<div><ol>
<li>What type of communities affected more from UHI in terms of age, sex, race, and poverty rate?
</li><div>
<li>Which regions has high UHI intensity?
</li>
<li>How vegetation index and digital elevation model affect UHI intensity?
</li>
</ol>


## Data description {.smaller .build}
<div>
- **Urban Heat Island data**
  + UHI = urban heat island (55,871 obs.)
  + also use NDVI (vegetation index), LST (land surface temperature); DEM (digital elevation model)
  + source: [Chakraborty, 2020](https://data.mendeley.com/datasets/x9mv4krnm2/3) (data.mendeley.com)
</div>
<div>
- **Demographic data**
  + age, sex, race at the census tract-level demographic data (74,001 obs)
  + source: [American Community Survey, 2017](https://data.census.gov/table?q=age+&g=0100000US$1400000&y=2017&tid=ACSST5Y2017.S0101) (data.census.gov)
</div>
<div>
- **Socioeconomic data**
  + poverty rate (74,001 obs); source: [American Community Survey, 2017](https://data.census.gov/table?q=age+&g=0100000US$1400000&y=2017&tid=ACSST5Y2017.S0101) (data.census.gov)
</div>
<div>
**Representative population:**

- Unit of analysis: tracts by census id (42,183 obs.)

- Population: all tract across US observed in 2017
</div>


## Descriptive Findings {.smaller .build}
Understanding relationship: UHI ~ Poverty rate
```{r preprocessing 1}
boxplot_uhi_highpov <- uhindvitract %>%
  select(UHI_summer_day, highpov) %>%
  mutate(type = highpov)
boxplot_uhi_shareblack <- uhindvitract %>%
  select(UHI_summer_day, shareblack) %>%
  mutate(type = shareblack)
boxplot_uhi_sharefemale <- uhindvitract %>%
  select(UHI_summer_day, sharefemale) %>%
  mutate(type = sharefemale)
boxplot_uhi_over64 <- uhindvitract %>%
  select(UHI_summer_day, shareover64) %>%
  mutate(type = shareover64)

boxplot1 <- ggplot(boxplot_uhi_highpov) +
  geom_boxplot(aes(x=highpov, y = UHI_summer_day)) +
  ggtitle('UHI - Poverty rate')
boxplot2 <- ggplot(boxplot_uhi_shareblack) +
  geom_boxplot(aes(x=shareblack, y = UHI_summer_day)) +
  ggtitle('UHI - Black share')
boxplot3 <- ggplot(boxplot_uhi_sharefemale) +
  geom_boxplot(aes(x=sharefemale, y = UHI_summer_day)) +
  ggtitle('UHI - Female share')
boxplot4 <- ggplot(boxplot_uhi_over64) +
  geom_boxplot(aes(x=shareover64, y = UHI_summer_day)) +
  ggtitle('UHI - Over 64 years share')
```

```{r output}
boxplot1+boxplot2+boxplot3+boxplot4
```

## Descriptive Findings {.verysmaller .build}
The top 20 cities with highest UHI are **low populated**, have **moderated female share**, and **high poverty**.
```{r preprocessing 2}
#top 20 cities in terms of urban heat island intensity
#cleanig urban data
uhindviurban <- uhindvitract %>%
  mutate(GEOID10 = Urban_geoid) %>%
  mutate(GEOID10 = as.character(GEOID10)) %>%
  mutate(GEOID10 = ifelse(nchar(GEOID10) == 3,
                          gsub("^(.{0})(.*)$","\\100\\2", GEOID10),
                          GEOID10)) %>%
  mutate(GEOID10 = ifelse(nchar(GEOID10) == 4,
                          gsub("^(.{0})(.*)$","\\10\\2", GEOID10),
                          GEOID10))

uhindviurban_picked <- uhindviurban %>%
  select(Urban_name, Urban_name_extended, NDVI_urb_summer, total,
         UHI_summer_day, DEM_urb, povt_rate, nblack, nfemale, nover64, GEOID10,
         white, black_or_african_american, asian_nativehawaiaan_pacific_islander, other)

urban_data <- uhindviurban_picked %>%
  group_by(GEOID10, Urban_name, Urban_name_extended) %>%
  summarise(total = sum(total),
            white = sum(white),
            black = sum(black_or_african_american),
            asian_pacificislander = sum(asian_nativehawaiaan_pacific_islander),
            other = sum(other),
            UHI_mean = round(mean(UHI_summer_day),2),
            NDVI_mean = round(mean(NDVI_urb_summer),2),
            DEM_mean = round(mean(DEM_urb),2),
            povt_rate = round(mean(povt_rate),2),
            shareblack = round(mean(nblack),2),
            sharefemale = round(mean(nfemale),2),
            shareover64 = round(mean(nover64),2)) %>%
  arrange(desc(UHI_mean))

urban_data_20 <- urban_data %>%
  ungroup() %>%
  slice(1:20) %>%
  select(Urban_name, total, UHI_mean, NDVI_mean,
         DEM_mean, povt_rate, shareblack, sharefemale, shareover64)
```

```{r output2}
urban_data_20 %>%
  kbl(caption = "20 cities with highest UHI intensity") %>%
  kable_classic_2(full_width = F) %>%
  kable_styling(font_size = 12,latex_options = "scale_down")
```

## Descriptive Findings {.verysmaller .build}
There is an **increase of UHI intensity** with the **increase of poverty rate**, and is statistically significant.

<div style="float: left; width: 50%;">
``` {r, echo = FALSE, results = 'markup', eval = TRUE, out.height=300, out.width=400}
ggplot(data = uhindvitract,
       aes(x=povt_rate, y=UHI_summer_day)) +
  geom_point(aes(size = total)) + #aes(size=total) if we want to weight it
  geom_smooth(method = "lm", formula = y ~ x + x^2) +
  ggtitle('Scatterplot between exposure to UHI and poverty rate') +
  labs(x="poverty rate", y="UHI intensity")                                    
```
</div>

<div style="float: right; width: 50%;font-size: 60%;">
```{r, echo = TRUE, results = 'markup', eval = TRUE, font-size: 5}
ols1l <- lm(UHI_summer_day ~ povt_rate,
            data = uhindvitract, weight = total)
summary(ols1l)
```
</div>


## Descriptive Findings {.verysmaller .build}
Majority black with higher poverty experienced **higher UHI exposure**.


<div style="float: left; width: 50%; font-size: 60%;">
``` {r, echo = FALSE, results = 'markup', eval = TRUE, out.height=300, out.width=400}
ggplot(data = uhindvitract,
       aes(x=nblack, y=UHI_summer_day)) +
  geom_point(aes(size=total)) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2)) +
  ggtitle('Scatterplot between exposure to UHI and black share') +
  labs(x="black share", y="UHI intensity")
```
  + Moreover, there is an increase in the exposure of UHI with the increase in poverty rate in dominantly black population area
</div>

<div style="float: right; width: 50%; font-size: 60%;">
```{r preprocessing t1_uhi_twd}
t1_uhi_wtd <-
  tapply(uhindvitract$UHI_summer_day * uhindvitract$total,
         list(uhindvitract$highpov, uhindvitract$shareblack),
         sum) / 
  tapply(uhindvitract$total,
         list(uhindvitract$highpov, uhindvitract$shareblack),
         sum)

t1_uhi_wtd_df <- as.data.frame(t1_uhi_wtd)

t1_uhi_wtd_df <- t1_uhi_wtd_df %>%
  mutate(diff = t1_uhi_wtd_df[,2] - t1_uhi_wtd_df[,1])

new_row_t1 <- c(t1_uhi_wtd_df[2,1] - t1_uhi_wtd_df[1,1], t1_uhi_wtd_df[2,2] - t1_uhi_wtd_df[1,2], t1_uhi_wtd_df[2,3] - t1_uhi_wtd_df[1,3])
t1_uhi_wtd_df <- rbind(t1_uhi_wtd_df, new_row_t1)
rownames(t1_uhi_wtd_df) <- c("Not high poverty", "high poverty", "diff")
```

```{r \tiny, echo = TRUE, results = 'markup', eval = TRUE}
t1_uhi_wtd_df
```

```{r preprocessing ols_b_l}
#get separate data frames by predominantly Black stations to estimate separate models
ct_black <- uhindvitract %>% filter(shareblack == "Majority Black")
ct_nonblack <- uhindvitract %>% filter(shareblack == "Majority non-Black")

#shareblack == 1: linear model with tract observations
ols_b_l <- lm(UHI_summer_day ~ povt_rate, 
              data = ct_black, weights = total)
```

```{r output ols_b_l}
summary(ols_b_l)
```
</div>




